---
title: "Using property testing in R"
description: |
  A short description of the post.
author:
  - name: Etienne Bacher
    url: {}
date: 2024-09-28
output:
  distill::distill_article:
    self_contained: false
draft: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I had never heard of property testing until a few months ago when I started looking
at some pull requests in [polars]() (the Python implementation, not the R one)
where they use [hypothesis.io](). 

<example PR?>

I have contributed to a fair amount of R packages but I have never seen this type
of tests before: unit tests, plenty; snapshot tests, sometimes; but property-based
tests? Never. And at first, I didn't really see the point but I've had a couple
of situations recently where I thought it could be useful, so this post is here
to describe those situations and why property-based testing can be useful.

## What is property-based testing?

Most of the time, unit tests check that the function performs well on a few different
inputs: does it give the correct results? nice error messages? what about this
corner case?

Property-based testing is a way of testing where we give random inputs to the function we want to test and we want to ensure that no matter the inputs, the output will respect some properties. For example, we made a function to reverse the input, so if I pass `3, 1, 4`, it should return `4, 1, 3`^[Example taken from the Rust crate [`quickcheck`](https://github.com/BurntSushi/quickcheck).]. We can pass several inputs and see if the output is correctly reversed. But a more efficient way would be to check that our function respects a basic property, which is that *reversing the input twice should return the original input*:

```{r}
rev(rev(c(3, 1, 4)))
```

Therefore, property-based testing doesn't use hardcoded values to check the output but ensures that our functions respects a list of properties. 

<demo + explain shrinking>


Besides this basic example, where could this be useful?

## Ensuring that a package doesn't crash R

When working with compiled code (C++, Rust, etc.), it can happen that a bug makes the R session crash (== segfault == "bomb icon" in RStudio). This can be extremely annoying as we lose all data and computations that were stored in memory. When we work with compiled code, there's one property that our code should follow:

> Calling a function shouldn't lead to a segfault.

This happened to me a few months ago. I investigated a part of some code that used `igraph::cluster_fast_greedy()`. I know almost nothing about `igraph`, I was just playing around with arguments, and suddenly... crash. I reported this situation ([igraph#2459](https://github.com/igraph/igraph/issues/2459)), which was promptly fixed (thank you `igraph` devs!), but one sentence in the explanation caught my eye: "it is a rare use case to only want modularity but not membership, and avoiding membership calculation doesn't have any advantages."

I have no particular problem with this sentence or the rationale behind, it makes sense to prioritize fixes that affect a larger audience. But it got me thinking: could we try all combinations of inputs to see if it makes the session crash? We could use [parametric testing]() for this, but then again we need to hardcode at least some possible values for parameters. We could say that we start by testing only `TRUE`/`FALSE` values for all combinations of params, but what if the user passes a string?

I think this is a situation where property-based testing would be helpful: we know that no matter the input type, its length, and the value of other inputs, *the session shouldn't crash*.

## Ensuring that a package and its variants give the same results

I have spent some time working on [`tidypolars`](), a package that provides the same interface as the `tidyverse` but uses `polars` under the hood. This means that there should be the lowest amount of "surprises" for the user: the behavior of functions that are available in `tidypolars` should match the behavior of those in `tidyverse`. Once again this can be tedious to check. One example is the function `stringr::str_sub()`. For instance, we can start with basic examples, such as:

```{r}
stringr::str_sub(string = "foo", start = 1, end = 2)
```

Easy enough. But what happens if `string` is missing? Or if `start > end`? Or if `end` is negative? Or if `start` is negative *and* `end` is `NULL` *and* the length of `start` is greater than the length of `string`? Testing all of those is painful and it increases the risk of forgetting a corner case.

It is better here to use property-based testing: we don't need to check the value of the output of functions implemented in `tidyplars`, **we only need to check that they match the output of functions in `tidyverse`**.

<example>


## Conclusion

Property-based testing will not replace all kinds of tests, and is not necessarily appropriate in all contexts. Still, it can help uncover bugs, segfaults, and it adds more confidence in our code.
