---
title: Compare two languages in R Markdown documents
author: Etienne Bacher
date: '2020-12-03'
description: 'A few months ago, I tried to compare two languages in an article. Here are several ways to do it.'
output:
  distill::distill_article:
    toc: true
    toc_float: true
    highlight: pygments
    highlight_downlit: true
---

In one of [my first posts](https://www.etiennebacher.com/post/first-contact/first-contact/), I wanted to compare two languages (namely and R and Stata) by putting two chunks side-by-side. I asked how to do this [on StackOverflow](https://stackoverflow.com/questions/59511268/display-several-code-chunks-in-a-concise-way) and continued to dig this question occasionally. I have now a few more or less convincing solutions/alternatives, that I summarize here.


## Two code chunks side-by-side

It is possible to produce two columns in an HTML document (this is also possible for LaTeX files but the code is different). The following code...

````
:::: {style="display: grid; grid-template-columns: 70% 70%; grid-column-gap: 30px;"}

::: {}
```{r}`r ''`
head(mtcars)
```
:::

::: {}
```{r}`r ''`
head(anscombe)
```
:::

::::
````

... produces this output:

:::: {style="display: grid; grid-template-columns: 70% 70%; grid-column-gap: 30px;"}
::: {}
```{r}
head(mtcars)
```
:::
::: {}
```{r}
head(anscombe)
```
:::
::::


All of this is more detailed in [this section](https://bookdown.org/yihui/rmarkdown-cookbook/multi-column.html) of the R Markdown Cookbook.



## Create tabs with `{xaringanExtra}`

An alternative to side-by-side chunks is to create tabs. We lose the ability to compare directly two chunks, but we can put much more tabs than code chunks. To do so, we use [`{xaringanExtra}`](https://pkg.garrickadenbuie.com/xaringanExtra/), made by Garrick Aden-Buie. It is a great package that adds a lot of functionalities to R Markdown or `{xaringan}`.

To create tabs, we run `xaringanExtra::use_panelset()` first, and then we create the sections. Let's init the panelset:

```{r}
library(xaringanExtra)
# enable panelset
use_panelset()
```

Next, we can create several panels with `::::: {.panelset}` and `::: {.panel}`. Here's an example:

````
::::: {.panelset}

::: {.panel}
[mtcars]{.panel-name}
```{r}`r ''`
head(mtcars)
```
:::

::: {.panel}
[anscombe]{.panel-name}
```{r}`r ''`
head(anscombe)
```
:::

::::
````

::::: {.panelset}
::: {.panel}
[mtcars]{.panel-name}
```{r}
head(mtcars)
```
:::
::: {.panel}
[anscombe]{.panel-name}
```{r}
head(anscombe)
```
:::
:::: 



## Use `<details>`

Finally, it is also possible to create chunks that are hidden by default but can be expanded by the user. This is particularly useful if you want to provide the user a reference. For instance, if you're trying to teach a new language, it might be helpful to provide code that the user already knows,  without displaying it by default.


Here's an example. We create the "expanding zone" below with:

````
<details>
<summary> Stata </summary>
Here, I put some Stata code hidden so that the user can compare if necessary:
```stata `r ''`
regress y x
```
</details>
````


```{r, eval=FALSE}
# Here I'm teaching R
lm(mpg ~ drat, data = mtcars)
```

<details>
<summary> Stata </summary>
Here, I put some Stata code hidden so that the user can compare if necessary:
```stata
regress y x
```
</details>


<br>

That's all! To summarize:

* code chunks side-by-side
* tabs
* hidden chunks with `<details>`






