---
title: "Some notes about optimizing base R code"
description: |
  A short description of the post.
author:
  - name: Etienne Bacher
    url: {}
date: 2022-10-10
draft: true
output:
  distill::distill_article
---

Lately I've spent quite some time on packages that require (almost) only base R:

* `datawizard`, a package belonging to the `easystats` ecosystem, whose goal is to 
  provide tools for data wrangling and statistical transformations;
* `poorman`, whose goal is to reproduce `tidyverse` functions (with a strong focus
  on `dplyr`) using base R only.
  
I've used `bench::mark()` and `profvis::profvis()` a lot to improve code performance 
and here are a few things I learnt. By default, `bench::mark()` checks that all
expressions return the same output, so we can be confident that the alternatives
I show in this post are truly equivalent.


Note that these performance improvements are targeted to package developers. A
random user shouldn't really care if a function takes 200 milliseconds less to
run. On the other side, I think a package developer might find these tips
interesting.

Before we get started, here's a small function that I use to make a classic dataset
(like `iris` or `mtcars`) much bigger.

```{r}
make_big <- function(data, nrep = 500000) {
  tmp <- list()
  for (i in 1:nrep) {
    tmp[[i]] <- data
  }
  
  data.table::rbindlist(tmp) |> 
    as.data.frame()
}
```



## Check if a vector has a single value

One easy way to do this is to run `length(unique(x)) == 1`, which basically means
that first we have to collect all unique values and then count them. This can be 
quite inefficient: it would be enough to stop as soon as we find two different 
values.

What we can do is to compare all values to the first value of the vector. Below is 
an example with a vector containing 1 million values. In the first case, it only
contains `1`, and in the second case it contains `1`s and `2`s.

```{r}
# Should be TRUE
test <- rep(1, 1e7)

bench::mark(
  length(unique(test)) == 1,
  all(test == test[1]),
  iterations = 10
)

# Should be FALSE
test2 <- rep(c(1, 2), 1e7)

bench::mark(
  length(unique(test2)) == 1,
  all(test2 == test2[1]),
  iterations = 10
)
```

This is also faster for character vectors:

```{r}
# Should be FALSE
test3 <- rep(c("a", "b"), 1e7)

bench::mark(
  length(unique(test3)) == 1,
  all(test3 == test3[1]),
  iterations = 10
)
```


## Concatenate columns

Sometimes we need to concatenate columns, for example if we want to create a 
unique id from several grouping columns.

```{r}
test <- data.frame(
  origin = c("A", "B", "C"),
  destination = c("Z", "Y", "X"),
  value = 1:3
)

test <- make_big(test)
```

One option to do this is to combine `paste()` and `apply()` using `MARGIN = 1` to
apply `paste()` to each row. However, a faster way to do this is to use `do.call()`
instead of `apply()`:

```{r}
bench::mark(
  apply = apply(test[, c("origin", "destination")], 1, paste, collapse = "_"),
  do.call = do.call(paste, c(test[, c("origin", "destination")], sep = "_"))
)
```



## Get row names

```{r}
# with numeric rownames
bench::mark(
  as.numeric(row.names(test)),
  attributes(test)$row.names
)

# with character rownames
row.names(test) <- as.character(1:1500000)

bench::mark(
  row.names(test),
  attributes(test)$row.names
)
```


## Giving attributes to large dataframes

This one comes from these [StackOverflow question and answer](https://stackoverflow.com/questions/74029805/why-does-adding-attributes-to-a-dataframe-take-longer-with-large-dataframes). Manipulating a dataframe can remove some attributes. For example, if I give an
attribute `foo` to a large dataframe:

```{r}
orig <- data.frame(x1 = rep(1, 1e7), x2 = rep(2, 1e7))
attr(orig, "foo") <- TRUE
attr(orig, "foo")
```

If I reorder the columns, this attribute disappears:

```{r}
new <- orig[, c(2, 1)]
attr(new, "foo")
```

We can put it back with:

```{r}
attributes(new) <- utils::modifyList(attributes(orig), attributes(new))
attr(new, "foo")
```

But this takes some time because we also copy the 10M row names of the 
dataset. Therefore, one option is to create a custom function that only copies the
attributes that were in `orig` but are not in `new` (in this case, only attribute
`foo` is concerned):

```{r}
replace_attrs <- function(obj, new_attrs) {
  for(nm in setdiff(names(new_attrs), names(attributes(data.frame())))) {
    attr(obj, which = nm) <- new_attrs[[nm]]
  }
  return(obj)
}

bench::mark(
  old = {
    attributes(new) <- utils::modifyList(attributes(orig), attributes(new))
    head(new)
  },
  new = {
    new <- replace_attrs(new, attributes(orig))
    head(new)
  }
)
```


